\unnumberedchapter{Declaration of Original and Sole Authorship} 
\chapter*{Declaration of Original and Sole Authorship} 

I, \name, declare that this thesis entitled \emph{\thesistitle} and the data presented in it are original and my own work. 


I confirm that:
\begin{itemize}
\item No part of this work has previously been submitted for a degree at this or any other university.
\item References to the work of others have been clearly acknowledged. Quotations from the work of others have been clearly indicated, and attributed to them.
\item In cases where others have contributed to part of this work, such contribution has been clearly acknowledged and distinguished from my own work.
\item None of this work has been previously published elsewhere, with the exception of the following: 
\begin{itemize}
\item Chapter~\ref{ch:1d} is largely derived from a paper published in the New Journal of Physics (\textbf{18}(3):035012, 2016)~\cite{schloss2016}.
\item Chapter~\ref{ch:2d} is largely derived from a paper published in Phys. Rev. Fluids (\textbf{4}(5):054701, 2019)~\cite{zhang2019}.
\item The GPUE codebase, which is a primary topic in Chapter~\ref{ch:gpu} has been published in the Journal of Open Source Software (\textbf{3}(32):1037, 2018)~\cite{schloss2018}.
\item Chapter~\ref{ch:vortex_states} is largely derived from a recently submitted manuscript to Phys. Rev. Fluids (arXiv:1910.02364)~\cite{schloss2019}.
In addition, many figures for this chapter have been created for the GPUE documentation~\cite{docs}.
\item Figure~\ref{fig:bench} was created by Peter Wittek when comparing different numerical solvers~\cite{wittek2016}.
\item Figure~\ref{fig:expr_tree} was largely derived from tikz code provided by Xadisten during a Twitch livestream.
\item Figure~\ref{fig:coalesce} was largely derived from tikz code created by user Realz Slaw on stackexchange~\cite{stackoverflow}.
\item Figure~\ref{fig:mode_plot} was Reproduced by Nieddu \textit{et al.} and Kumar \textit{et al.}~\cite{nieddu2016, kumar2015}.
\end{itemize}
All of the articles mentioned above have been (or will be published) under Creative Commons BY with attribution to the original authors, and in each chapter, I clearly define my focus in each publication.
\end{itemize}

Date:  \submissiondate

Signature: 




